Flaring star-planet interactions in Kepler and TESS

### 2023-06-02 src/tex/ms.tex
[showyourwork] overleaf sync

If we picture the magnetic field lines connecting the planet to the star in the sub-Alfv\'enic zone, we expect that the stellar footpoint of these lines will move along with the planet. Flares triggered by the planet will occur at a preferred orbital phase, namely whenever the footpoint is visible to the observer. Depending on the system's viewing geometry, i.e., the inclination of the rotation axis, the spin-orbit (mis-)alignment, and the extent and configuration of the large scale magnetic field -- the footpoint could be always in view, or never. In expectation, however, in star-planet systems with favorable viewing geometries and magnetic field strengths and configurations, the footpoint visibility will be modulated in phase with the planetary orbit~(see, e.g., Fig.~4 in \citet{strugarek2019chasing} for an illustration XXXX maybe make my own sketch? XXXX). The resulting phase preference of planet-induced flares introduces a deviation from a random distribution of flares. Space missions like Kepler and TESS provide light curves that cover all phases of the planet's orbit, so that phase-correlated planet-induced flares can be detected against a background of uncorrelated intrinsic flaring.

@@

The average Alfv\'en surface size ($\overline{AS}_{R}$) was estimated using the power law  (Eq.~\ref{AS_estimation}) established by Chebly et al. (under revision):

@@

This relation links the mean AS size with the maximum large-scale radial magnetic field ($B_{\rm r}^{\rm max}$) obtained in Zeeman-Doppler Imaging reconstructions  \textcolor{red}{(references here)}. The relation was derived from 3D~MHD stellar wind simulations cool main sequence stars (F, G, K, and M), which incorporated observed ZDI maps as boundary condition for the stellar radial magnetic field (for more details on the models see Chebly et al. 2023). However, the stellar magnetic field~($B$) in this work (Table~\ref{tab:maintable_der}), derived from the Rossby number of each star (Eq.~\ref{Eq3}, \ref{Eq4}), corresponds to the total, unsigned magnetic field strength on the stellar surface. As such, it is necessary to estimate how $B_{\rm r}^{\rm max}$ could be related to $B$.

@@

\subsection{Intermittency}
\label{sec:discussion:intermittency}
Magnetic star-planet interaction power in a given system may sometimes be low for episodes of time, and high for others. HD 179949 is the prototype example for this phenomenon. The orbital modulation of its chromospheric indicators was observable for four out of six epochs on a five-year baseline~\citep{shkolnik2003evidence,shkolnik2008nature}. It is not clear how long a continuous epoch of interaction or non-interaction might be. This leaves room for many different explanations. Uninterrupted, long-term monitoring of individual systems is required to investigate the reasons for the temporary cessation of interaction. Candidate mechanisms for the star include: 
\begin{itemize}
    \item the Alfv\'en radius moving in- and out the planet's orbit, either over the course of an activity cycle, or through short-term variations in the magnetic field and wind properties caused by, e.g., coronal mass ejections;
    \item the interacting footpoint moving to higher latitudes due to changes in the large scale field, which reduces the modulation of its visibility 
    \item an active latitude (i.e., small scale field) moving away from the magnetic footpoints passageway; or
    \item the available magnetic energy decreasing such that the flares or other magnetic interaction indicators still occur but fall below the detection threshold for a period of time.
\end{itemize}

@@

The first model measures the torque $\partial L_{\rm conv} / \partial t$, exerted by angular momentum transfer between star and planet, assuming the star has a convective envelope~(\citealt{penev2012constraining}, their Eqns. 1 and 2). We choose the tidal quality factor $Q_*=10^7$. Our results in Fig~\ref{fig:tidal} (left panel) do not depend on the choice of $Q_*$, as we only need to know the tidal torque down to a proportionality factor. We also make the simplifying assumption that the rotation period of the convective envelope is the same as the surface rotation period measured as described in Section~\ref{sec:data:rotationperiods}, ignoring differential rotation effects.

>>>>

If we picture the magnetic field lines connecting the planet to the star in the sub-Alfv\'enic zone, we expect that the stellar footpoint of these lines will move along with the planet. Flares triggered by the planet will occur at a preferred orbital phase, namely whenever the footpoint is visible to the observer. Depending on the system's viewing geometry, i.e., the inclination of the rotation axis, the spin-orbit (mis-)alignment, and the extent and configuration of the large scale magnetic field -- the footpoint could be always in view, or never. In expectation, however, in star-planet systems with favorable viewing geometries and magnetic field strengths and configurations, the footpoint visibility will be modulated in phase with the planetary orbit~(Fig.~\ref{}). The resulting phase preference of planet-induced flares introduces a deviation from a random distribution of flares. Space missions like Kepler and TESS provide light curves that cover all phases of the planet's orbit, so that phase-correlated planet-induced flares can be detected against a background of uncorrelated intrinsic flaring.

\begin{figure*}[ht!]
\script{paper_illustrative_sketch_of_system_geometries.py}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/system_geometries.png}
        \caption{
         Simplified magnetic star-planet interaction geometries.\textbf{ Panel a: Fully aligned system along the line of sight. }Magnetic, rotational, and orbital axis are aligned. The interaction footpoint moves along the black line, and is visible on the red line, i.e. during $50\%$ of the planet's orbit. Flares induced along this line will result in excess flares when the planet is in front of the star. \textbf{Panel b: Orbital misalignment.} The rotational and magnetic axis are aligned, but the planet is highly misaligned. The footpoint of interaction moves between the two horizontal black lines, and is still visible $50\%$ of the time, resulting in modulated flaring with orbital phase. However, the footpoint passes different latitudes on the star, which correspond to different dipole field strengths, which may influence the power of interaction. \textbf{Panel c: Inclined fully aligned system.} The red interaction footpoint line is visible $100\%$ of the time. No modulation due to SPI can observed, even if flares are triggered by the planet. The sketch only shows the interaction footpoint region on the Northern hemisphere of the star, but they may appear on both.
        }
        \label{fig:sketch}
    \end{centering}
\end{figure*}

@@

The average Alfv\'en surface size ($\overline{AS}_{R}$) was estimated using the power law  (Eq.~\ref{AS_estimation}) established by \citet{Chebly2023}:

@@

This relation links the mean AS size with the maximum large-scale radial magnetic field ($B_{\rm r}^{\rm max}$) obtained in Zeeman-Doppler Imaging reconstructions (\citealt{donati2006surprising, morin2008largescale, fares2009magnetic, alvarado-gomez2015activity, hussain2016spectropolarimetric, kochukhov2020mapping}). The relation was derived from 3D~MHD stellar wind simulations cool main sequence stars (F, G, K, and M), which incorporated observed ZDI maps as boundary condition for the stellar radial magnetic field (for more details on the models see \citealt{Chebly2023}). However, the stellar magnetic field~($B$) in this work (Table~\ref{tab:maintable_der}), derived from the Rossby number of each star (Eq.~\ref{Eq3}, \ref{Eq4}), corresponds to the total, unsigned magnetic field strength on the stellar surface. As such, it is necessary to estimate how $B_{\rm r}^{\rm max}$ could be related to $B$.

@@

\subsection{Intermittency}
\label{sec:discussion:intermittency}
Magnetic star-planet interaction power in a given system may sometimes be low for episodes of time, and high for others. HD 179949 is the prototype example for this phenomenon. The orbital modulation of its chromospheric indicators was observable for four out of six epochs on a five-year baseline~\citep{shkolnik2003evidence,shkolnik2008nature}. It is not clear how long a continuous epoch of interaction or non-interaction might be. This leaves room for many different explanations. Uninterrupted, long-term monitoring of individual systems is required to investigate the reasons for the temporary cessation of interaction. Candidate mechanisms for the star include: 
\begin{itemize}
    \item the Alfv\'en radius moving in- and out the planet's orbit, either over the course of an activity cycle, or through short-term variations in the magnetic field and wind properties caused by, e.g., coronal mass ejections;
    \item the interacting footpoint moving to higher latitudes due to changes in the large scale field, which reduces the modulation of its visibility 
    \item an active latitude (i.e., small scale field) moving away from the magnetic footpoints passageway; or
    \item the available magnetic energy decreasing such that the flares or other magnetic interaction indicators still occur but fall below the detection threshold for a period of time.
    \item the availability of a sufficient amount of plasma in the vicinity of the magnetic interaction region to yield significant observable flare emission in the stellar atmosphere (in analogy to the Jupiter-Io system, wherein Io's volcanism provides the plasma that eventually produces an interaction hotspot on Jupiter,~\citealt{clarke1996farultraviolet, prange1996rapid})
\end{itemize}

@@

The first model measures the torque $\partial L_{\rm conv} / \partial t$, exerted by angular momentum transfer between star and planet, assuming the star has a convective envelope~(\citealt{penev2012constraining}, their Eqns. 1 and 2). We choose the tidal quality factor $Q_*=10^7$, i.e., low efficiency of tidal dissipation, as a conservative estimate. Our results in Fig~\ref{fig:tidal} (left panel) do not depend on the choice of $Q_*$, as we only need to know the tidal torque down to a proportionality factor. We also make the simplifying assumption that the rotation period of the convective envelope is the same as the surface rotation period measured as described in Section~\ref{sec:data:rotationperiods}, ignoring differential rotation effects.
